Cover letter for the original IJRR submission

### 2017-09-28 reviews_tro/reviewer_response.tex

\documentclass[a4paper]{article}
 \usepackage{fullpage}
\usepackage{amsfonts} % pour les lettres maths creuses \mathbb
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[frenchb]{babel}
\usepackage{aeguill}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{pdfpages}
% Another method to track changes
% Examples of usage:
% - This is \added[id=per,remark={we need this}]{new} text.
% - This is \deleted[id=per,remark=obsolete]{unnecessary}text.
% - This is \replaced[id=per]{nice}{bad} text.
% To print the list of change use: \listofchanges
\usepackage{changes}	% use this for the working version
%\usepackage[final]{changes} % use this for the final version
\definechangesauthor[name={Andrea Del Prete}, color=orange]{adp}
\newcommand{\deladp}[1]{\deleted[id=adp]{#1}}
\newcommand{\addadp}[1]{\added[id=adp]{#1}}
\newcommand{\repadp}[2]{\replaced[id=adp]{#1}{#2}}
\definechangesauthor[name={Steve Tonneau}, color=blue]{st}
\newcommand{\delst}[1]{\deleted[id=st]{#1}}
\newcommand{\addst}[1]{\added[id=st]{#1}}
\newcommand{\repst}[2]{\replaced[id=st]{#1}{#2}}
\newcommand{\gls}[1]{\textit{#1}}
\newcommand{\glslink}[2]{{#2}}
\newcommand{\done}[0]{\textcolor{green}{DONE}}
\newcommand{\ndone}[0]{\textcolor{red}{TODO}}

@@

\newcommand\dx{\dot{x}}
\newcommand\dy{\dot{y}}
\newcommand\ddx{\ddot{x}}
\newcommand\ddy{\ddot{y}}
\def\real{{\mathbb R} }
\newcommand\quot[1]{\begin{quote} \textit{quote}: \textbf{#1}\end{quote}}
\newcommand\as[1]{\begin{quote} \textit{answer}: {#1}\end{quote} \leavevmode \\ }
\frenchbsetup%
{%
StandardItemLabels=true,%
ItemLabels=\ding{43},%
}%
\DeclareUnicodeCharacter{00A0}{ }
\author {}
\title {Cover letter for the resubmission of our paper ``An efficient acyclic contact planner for multiped robots''}
\date {}
\begin{document}
\maketitle

\section{Answers to reviewer 1}

\quot {As for novel contributions, this did not seem to be addressed/explained
until page 3 (in the last paragraph of Sec. I). Much of the methodology
has already been presented in earlier work, and the particular aspects
that set THIS paper apart seem (as claimed by the authors) to be: 
}

\as{We agree that our original submission is not clear on the novelty between this paper and previous contributions.
This paper must be understood as an extension of the ISRR paper [20], while the other cited contributions from our team 
are exploited to validate our approach in a complete pipeline.
In this extension, the algorithm used to compute a sequence of contacts is the same as in [20] (although this
algorithm is not explicitely given in [20]), with the introduction of a fastest test for verifying the robust quasi static equilibrium of a contact posture.
We reformulated the end of Section I.B to state clearly that the novelty of this paper essentially lies in the experimental setup and validation we achieved with real robot models, in particular
in the empirical justification of the method, and its validation in a dynamic simulation. This effort led to the proposal of the first open source, complete framework for multi contact motion synthesis, since 
the validation required to write a solver for the problem P3. }\done

@@

\quot { (1) A criterion for robust static equilibrium, but it isn't
described until page 12, in Appendix B, Section B. [\dots]  It would really help to describe this MUCH earlier in
the paper (than page 13), if it is a significant novelty of your paper.}

\as{We agree with the remark. To better distinguish between what is novel and not, we modified the paper to move the content of Appendix B.B (now Section VI), as well as D  (the pseudo code of our algorithm, now Section V.B.2) and E (reference to our source code, now Section VII) to the text itself. This results in the criterion being presented much earlier.}\done

@@

\quot{It is a pretty intuitive/typical check for quantifying a
margin for friction conesque violations (using an approximation for
the friction cone, with x and y direction forces considered
independently). }

\as{Thanks for the comment. To our knowledge, what is common in the literature is to reduce the friction coefficient to obtain more conservative bounds on the admissible forces, and thus a ``safety margin'' to the real cone boundary. An issue with this formulation is that the resulting margin is really small near the origin of the cone, and really important as we go further away from it. With our formulation, the obtained margin is constant everywhere on the cone, and thus guaranteed to be greater than the value $b_0$ we compute. We believe this formulation is new, but would be happy to cite any relevant reference if we happen to be wrong. This formulation proves useful to us because the margin $b_0$ allows to choose the more robust candidate for static equilibrium.}\done

@@

\quot{
Provides complete pseudocode.  This is great! However,it's
not fully clear how much of this is novel vs just re-presenting the
same algorithms previously publilshed.	If only code in App. B and C
are new, that should be clear. Conversely, if most of this presentation
was NOT outlined in such detail before, that should also be clear (to
the authors' credit). Either way, a brief clarification in the intro is
recommended.}

\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel at the end
of Section I.B.}\done

@@

\quot{(3)	The contribution to provide a solution to P3 doesn't happen
until page 13 (Appendix C) and as with contribution (1), it needs a
better, intuitive summary (briefly) when first mentioned, as well as a
roadmap for the reader as to where it will be described. It is also
ambiguous in the paper whether to what extent (if any) this solution is
novel, versus a presentation of work from [1]. "Provide a solution"
seemed to imply novelty, in the context of this paragraph -- yet later,
the authors (near the end of page 2) state they use a "state-of-the-art
solution to P3", which implies the P3 solution is not a claim of
novelty.}

\as{Indeed, we were not clear on P3. In the new version of the paper, we do not consider that our solution
is a new contribution, although our formulation differs from [1] (our control variable is the acceleration
of the center of mass, where in [1] the control variable is the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
was required to validate our approach. The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2.
We rephrased the text to clearly indicate that the algorithm is not a contribution. We made the choice not to give an intuition nor describe the solver in the core
of the paper because it is not required to understand the contributions to P1 and P2.}\done

@@

\quot{
(4)	The last contribution is validation using an in-house
simulator, as opposed to more approximate (unrealistic avatar) models
used in previous work.	As with the previous contributions claimed, you
should flesh out just a bit more detail, along with a roadmap to where
a particular contribution appears in the paper, to prepare and guide
the reader a bit more. More precisely, what are the differences between
these two simulation approaches (previous and current)? You need a
better description of what your in-house simulator does and does not
do; a vague citation to another paper based on that simulator [38] is
not sufficient.}

\as{Thanks for the comment. Regarding the absence of details on the dynamic simulator, first of all we apologize because [38] is not the appropriate reference.
We added a section where we describe the simulation in more details.

Regarding the difference between our previous and current work, the postures computed in [20] were never validated in a simulator, nor the complete motions since 
we did not have a solution to P3 at the time. Then the main
difference between the avatars of [20] and the real robots we now address are the joint limitations and auto-collisions probabilities.
The robots are much more constrained in this regard, which makes the planning harder. We thus had to demonstrate that our algorithm is also working
for these models, while maintaining decent performances.
We added this missing piece of information to the paper, at the end of Section I.B.}
\textcolor{red}{TODO ANDREA}.

@@

\quot{[\dots] the authors need to better
distinguish what the novel contributions are, and to highlight where
these contributions will be presented in the paper.  Also, many
important details are in the appendices. The paper would be improved
with more intuitive comments/summaries on these details, rather than
just writing (for example), we use a bunch of heuristics, and they are
described in Appendix X.}

\as{We agree. The novel details of the paper have been moved from the appendix to the core of the paper, and are explicitely detailed at the end of Section I.B.} \done

@@

\quot{
Also, where you describe "sensitivity" vs "specificity", it would be
useful to give a more intuitive explanation, example [\dots]}

\as{We updated Section VIII.B.1) to illustrate the meaning of those terms.} \done

@@

\quot{
However, I suggest the paper will be improved if
the presentation focuses more on highlighting (a) what is novel, and
(b)
what take-away messages you have, about why the whole framework is
designed the way it is [within the main text; not hidden in
appendices].}

\as{TODO}\ndone

\quot{
p.6 – this would be a great place to give more ``intuition'' for the heuristics. Hw seems to be used in essentially all foot contact, with $h_{efort}$ for hand/arm contacts, and ALSO $h_{vel}$ not seeming to be used at all.}
\as{Thanks for the suggestion. The text was updated accordingly, and the heuristic $h_{vel}$ was removed since indeed it is not used.} \done

@@

\quot{
p.6 six fingered hand is slightly misleading, since no rolling (as fully admitted in the text). I'd just cut this? \\
p.9 Not sure if Fig. 14 is worth including. It looks like its just Fig. 9 from the ISRR paper (?) [20], and there is still no ``full solution'' here -
 just an illustration of potential future work. 
 (At a minimum, the figure/solution should probably be referenced as having already appeared in the previous ISSR paper?)
}

\as{As also suggested by reviewer TODO, we removed the references to the results of the previous work [20], including those.}\done

@@

\quot{
p.8– somehow, “Table II” doesn’t appear until after “Table IV” – which is a frustrating choice for latex to make. Not sure what you can do to change that, but it would be preferable to have tables (and figures) appear in the correct order.
}

\as{We reorganized the paper such that the Table appear in the right order.}\ndone

@@

\quot{
p.8. Need to define ``discretization step'' more precisely, I think(?)… Min distance? Max distance? Nominal distance? In time? In Distance? In all 6 DOF??
}

\as{We updated the text to indicate that the discretization step refers to the number of postures generated along a path. For instance, a discretization step of 0.1 means
that 10 postures will be generated along a path of length 1. Here the length is the distance traversed by the root along the path, computed as a ponderation between the 3D position and joint variations of the root.

Finally, we also took into account all of the remaining suggestions in the pdf file attached. We do not detail them here since they refer essentially to the style 
and the writing of the paper.}\ndone

@@

\quot{
p.9 ‘addresses highly constrained environment’ – actually, not very well, correct? This was the “exception” case, for which results were rather poor, right? Also a case where “heuristics” are not so good for getting true (reliable) holds on grasp points, etc.?
}

\as{\textbf{\textcolor{red}{This quote annoys me, and I kind of want to maintain our statement. Addressing the car scenario, even with a success rate of 68 \%, seems like a nice achievement to me.
Also, in my opinion all our scenarios are highly constrained. What do you guys think?}}}\ndone

\quot{
The first part of App. B is referred to within this Appendix as “new minor contributions derived from previous work”. These aren’t so novel, perhaps, and what is more useful to the reader is a better intuition for when to use each within the MAIN BODY of the paper, where they are originally described/associated to particular simulations.
}

\as{Agreed. We removed the incriminated text and provided an intuition of the heuristic in the text.}\ndone

@@

\quot{github link seems to be dead? (The link on p. 15 is OK… should they be the same?)}

\as{We are suprised by this comment because after testing it, the link seems to be valid. In any case, in accordance with reviewer TODO, we changed all the links source codes to provide a SHA link that will bring the user to the accurate revision of the projects considered.}\ndone

\quot{remaining remarks in the pdf file provided by reviewer 1}
\as{The remaining comments from reviewer 1 were all addressed. Because they did not call for a discussion, as they are essentially related to style or typos, we do not list them here.}\ndone

@@

\section{Answers to reviewer 4}

\quot{ I suggest that a few papers related to more general motion and contact planning for multi-legged robots can be added. Particularly, the problem of ``foothold selection'' for 4 or 6-legged robots travelling over uneven terrains is a problem related to the one tackled in the paper. It would be nice to select a few representative works (e.g. from the DARPA Learning Locomotion project or
papers related to the ETHZ StarlETH and ANYmal robots) and explain the differences between those approaches and the problem solved in the paper.}
\as{Thanks for the relevant comment. As observed by the reviewer, our state of the art voluntarily focuses on multi-contact planning related work, but it is relevant
to mention the mentioned papers. We added a paragraph at the end of the state of the art to discuss them. As a matter of fact, we believe these approaches could be considered within
our framework to replace specific parts.}\ndone

@@

\quot{in my opinion a T-RO regular paper should be more self-containing. In several points the authors heavily rely on previous publications, e.g. when presenting the RRT-based planner (Section IV-C). This part should be expanded, describing in more detail what exactly has been changed with respect to the reference paper. }
\as{Thanks for the comment. As it is also suggested by reviewer 1 and the AE, we modified the paper for it to be self contained. The references to previous work now solely indicate
that a detailed method is not a contribution of the present paper. \\

In the particular case of Section IV-C, in fact the code of our implementation of the Bi-RRT with respect to the original one is exactly the same, with a single difference: when
traditionally a collision detection predicate is called to validate a configuration, we replace this predicate with a call to the reachability condition. We modified the text to explain this more clearly. As the Bi-RRT is a famous standard algorithm in the literature, we do not think it is necessary to include it in this paper, but will gladly do so if required. }

\quot{I am a bit skeptical about the readability of the figures. Some of them seem to be too small (e.g. Fig. 3, Fig. 4), some are of quite poor quality (why the robot silhouettes in Fig. 4 are blurred ?), and the purpose and meaning of Fig. 6 is rather unclear.}
\as{Thanks for the comment, with which we agree in general.  \\We increased the size of Fig. 3 and 4. The robot silhouettes in Fig. 4 are blurred when they describe a previous state of the robots. We added this precision in the text. \\Regarding the quality of our Figures, we are quite suprised by the comment. All of our simulation Figures are taken from high resolution rendering in blender. The other Figures are drawn from manually designed powerpoint vectorial graphics (converted to png). If the reviewer would be so kind as to point out specifically which Figures are of poor quality, we will gadly replace them.\\ As proposed also by reviewer 1, we removed references to previous results obtained in [20], so the Fig. 6 has been removed.}\ndone

\quot{In general, the simulation results presented qualitatively in the paper are a bit hard to interpret without the accompanying video, particularly if someone looks at a b/w print of the paper (as it may be looking in T-RO).}
\as{\textbf{\textcolor{red}{TODO This is true, although I think he has a crappy printer because I don't think it's that bad. I don't know what to do about it though. I can't simply remove the Figures,
because he tends to say he would like the paper to be more independent from the video...}}}\ndone

@@

\quot{ I also found the description of the environment model being
used,
and the way it is used in planning quite superficial. It is said, that
``an octree data structure'' is used to keep the sampled the
end-effector positions, and it is ``intersect'' with the environment.
The idea is OK, I understand that the octree is a sample-based
representation of the effector's workspace, and using some environment
model it is possible to decide if the given configuration can provide a
contact point or not. but how is the environment described ? How
efficient is this intersection operation ? Does this method require the
environment to be described by some specific objects (e.g. polygons) ?
What about the (unavoidable) uncertainty if a world model acquired via
on-line sensing is considered ? Perhaps some of these issues (e.g.
uncertainty) are not in the scope of the paper, but I can imagine that
a robotic researcher who reads the paper and is about to use the open
source software it advertises is very interested in the answers.}

\as{Thanks for the comment. The choice of not developing too much on the environment description and octree goes along with 
our effort to maintain the paper as readable as possible, since it already contains various concepts. However,
we agree that additional information must be added. We added a Section TODO where we indicate that the environment is represented as a polygon soup (or 3D mesh), and
that we assume that it is fully known. Although the integration of on-line sensing and state estimation in the planner is an issue currently considered by our team, it is indeed
out of the scope of this paper. We thus also indicate that we assume the environment and state of the robot are fully known. We also indicated the algorithmic complexity of the intersection
method to describe how efficient the intersection operation is.} \ndone

\quot{I appreciate the thorough presentation of results in the paper, however
it seems that this presentation is somewhat biased towards simulation
results. Simulations are perfectly OK for planning algorithms, but if
you decided to present the experimental results (HRP-2), then at least some
illustrations should be included in the paper. now they are only in the
accompanying video.  also a comparison between the results of the same
task(s) executed in dynamic simulation and on the real robot should
convince the reader that the planning method's good performance does
not depend too much on the idealized simulation environment. }
\as{We agree with the remark, and included some illustrations of the experimental results on HRP-2 in the paper. \\
We now elaborate on the validity of the simulation with respect to an experiment on the real robot. As also suggested by reviewer 1, we added a paragraph
in the paper (TODO where) to provide more information on the simulation we use to validate our motions. This simulation is an implementation of the state of the art approaches used by the community. Of course, it does not guarantee that any motion it validates will be accurately played on the robot, but it is used extensively by the community as a reasonable guarantee of feasibility.  The suggestion of the reviewer to compare the simulation results with those on the real robot is interesting. However, our robot HRP-2 is not torque controlled, but position controlled, therefore as of today a comparison is not possible.}\ndone

\section{Answers to reviewer 6}
\quot{ My primary concern, however,
is that the elements described in this paper have been described in
previous published work, which makes the particular contributions of
this work unclear. To be specific, the reachability-guided planner
appears identical to the planner in [20] (as stated by the authors in
the end of Sect. I.B.), and the optimal control formulation is from [1]
(again as stated in I.B.).}
\as{Thanks for the comment. Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated. We updated the paper to reflect the following. Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20]. We believe that this paper is thus in accordance with the guidelines of T-RO regarding the extension of a conference paper.}\ndone

### 2017-09-28 reviews_tro/reviewer_response.tex

\quot{The authors highlight a particular
contribution of this work to be the efficient robust balance criterion,
but I had trouble understanding this particular contribution. Does this
refer to the approximation of equilibrium feasibility by contact
reachability? In that case, how does this differ from just applying the
planner in [20] directly? Or does this refer to the robustness
parameter $b_0$ in Appendix B Eq. (15)?}
\as{Indeed, this part was also not clear. The robust balance criterion (renamed equilibrium criterion) indeed refers to the parameter $b_0$ in Appendix B. The solved
LP replaces the static equilibrium test performed in [20], which was not computationally efficient, and took up to 4ms per candidate to compute. We updated the paper to explain
this in details.}\ndone

\quot{In addition, sections IV and V are very similar to the description of
the planner in [20]. That's not surprising, of course, but there is no
indication in those sections that the work being presented is a
rephrasing of an earlier publication.[\dots] }
\as{We agree with the comment, and indeed made this clearer in the new submission.}\ndone

\quot{I do, however,
believe that the fact that the technical approaches are rephrasings of
prior work should be made more clear (or that sections which can be
replaced by citations of [20] and [1] be removed).}
\as{Thanks for the advice. This suggestion was in contradiction with the other reviewers, who thought that the paper should be self-contained.
The associate editor decided that the paper must be self contained, so we maintained the indicated sections, but made clearer the fact that they are rephrasing of previous work. }\ndone

\quot{*I think I understand what the authors mean by solving at
"interactive" rates, but this is worth defining. In my mind, it means
something like "before the human operator gets bored", or roughly > 0.5
Hz. But since the authors make the claim that this is the "first
interactive implementation of a contact planner", then that statement
should be made quantitative and verifiable}
\as{We agree with the relevant comment, and updated the paper accordingly.}\ndone

\quot{* In Section I.A. the authors refer to Bretl's work on multi-contact
planning as the inspiration for the P1,P2,P3 decomposition of the
problem. But it is not, as the authors state, always necessary to
decompose motion planning into choosing a series of quasi-static
configurations. There are a variety of feasible motions with no static
equilibrium configurations, so it is worth mentioning that this
decomposition only solves a subset of motion planning problem.}
\as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
is always an option). We updated the state of the art to indicate that the decomposition in general is restricitve.}\ndone

\quot{* Also in Sect. I.A, the authors state that "optimization-based
approaches only converge locally". This is not necessarily true, as
convex or mixed-integer convex optimizations can (sometimes) be solved
to a verifiable global optimum. I would suggest removing this sentence
entirely.}
\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene make the problem highly non-convex, and
that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima. We followed the reviewer's advice and simply removed the sentence. }\ndone

\quot{* I am not sure why the $C_{reach}$ hypothesis is described as the ``strong''
hypothesis in, e.g., Sect. IV. Is there a corresponding weak
hypothesis?}
\as{We apologize for the confusion, that comes from a inaccurate use of English by the main author. What we meant was that the hypothesis is TODO}\ndone

\quot{* In Sect. IV.B, the authors state that the ideal shape $B^*$ ``has no
explicit definition''. For a particular configuration  $q_0$, $B^*$ is simply
the surface of the robot, right? So I assume that this claim refers to
there being no explicit definition of B* which is the volume around
$B^*(q_0)$ for every $q_0$? I think I agree with the authors, but I found
this claim hard to follow and would appreciate some clarification}
\as{We updated the Section IV B. to better explain our claim. Indeed, as the reviewer states $B^*$  is the volume for every $q_0$. If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
We of course doubt that $B^*$ actually exists. }\ndone

\quot{* In Sect. V.A., the authors mention that the ability to plan a contact
break and creation with the same limb allows the creation of dynamic
motions. This is quite an important ability for an otherwise
quasi-static planner, but this ability isn't described again until the
appendix. It would be worth explaining why this is true in the paper
itself. }
\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}\ndone

@@

\quot{In Sect. V.B, the authors mention inserting an additional state
between contact configurations if two or more contacts cannot be
maintained. Shouldn't this step appear somewhere in Alg. 1?}
\as{We apologize for the lack of clarity. This is actually handled by the method \textsc{RepositionContacts}. We now realize that the name is misleading. Indeed,
\textsc{RepositionContacts} either adds a new contact, or tries to reposition one already existing. In Appendix D, we state:
\textit{Otherwise, the method \textsc{RepositionContacts} is called. It repositions one end effector (either a free
limb, or the oldest active contact) towards a new contact
position if possible. }

We changed the name to \textsc{TODO}, with the hope that it makes things clearer.}\ndone

\quot{* In Appendix A.A, it is mentioned that step 2 in the workspace
computation is conservative, with reference to figure 17-right. But
Fig. 17-right is the simplification of the convex hull, corresponding
to step 3, and it's actually hard to tell from the figure that it is
conservative. Not knowing anything about Blender's decimate tool, I
assume it is operating by removing vertices, which would indeed be
conservative, but some clarification would be useful here.}
\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}\ndone

\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
of the github links. I've personally made the mistake of linking to a
github project in a paper, only to have that project change and
eventually delete the specific code I was citing. }
\as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project.
Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}\ndone

### 2017-09-28 body/abstract.tex
ongoing, added TODO / DONE macros in letter

The reduction breaks the algorithm complexity encountered
in previous works, resulting in the first interactive
implementation of a contact planner (open source). While
no contact planner has yet been proposed with theoretical
completeness, we empirically show the interest of our framework:
in a
few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand. These plans are validated either in dynamic simulations, or on the real HRP-2 robot.

>>>>

The reduction breaks the algorithm complexity encountered
in previous works, resulting in the first interactive
implementation of a contact planner (open source). While
no contact planner has yet been proposed with theoretical
completeness, we empirically show the interest of our framework:
in a
few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ. These plans are validated either in dynamic simulations, or on the real HRP-2 robot.

### 2017-09-28 body/appendix_algo.tex
ongoing, added TODO / DONE macros in letter

\section{Algorithms for the discretization of a path}
\label{app:contact}

@@

\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}. It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

@@

>>>>

\subsubsection{Pseudo-code of the Algorithm}
\label{app:contact}

@@

@@

\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
 It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

### 2017-09-28 body/conclusion.tex
ongoing, added TODO / DONE macros in letter

\subsection{Addressing $\mathcal{P}_3$}
This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation,
. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
collision avoidance.
%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}

>>>>

\subsection{Addressing $\mathcal{P}_3$}
This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation
. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
collision avoidance.
%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}

### 2017-09-28 body/contact.tex
ongoing, added TODO / DONE macros in letter

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.

First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item At most one contact is added between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

@@

\subsubsection{Maintaining a contact in the sequence}

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts}
Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest. If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others.
%~ \begin{enumerate}
%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
%~ \end{enumerate}

@@

@@

Finally, to verify that a configuration is in static equilibrium, we use a new robust LP formulation. It replaces the computationally inefficient double description
approach used in our previous work~\cite{tonneauisrr15}.

>>>>

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
We first give an intuition of the algorithm, before providing its complete pseudo-code.
%~ The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
\subsubsection{Algorithm overview}
First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item At most one contact is added between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

@@

\paragraph{Maintaining a contact in the sequence}

@@

%\subsection{Generation of a contact configuration}  
\paragraph{Creating contacts}
Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest. If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others. \\ \\
%~ \begin{enumerate}
%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
%~ \end{enumerate}

@@

\input{body/appendix_algo}

@@

Finally, to verify that a configuration is in static equilibrium, we use a new robust LP formulation. It replaces the computationally inefficient double description
approach used in our previous work~\cite{tonneauisrr15}, and presented in the following Section~\ref{sec:Equil}.

### 2017-09-28 body/efort.tex
ongoing, added TODO / DONE macros in letter

%~ \section{Heuristics and criteria for contact selection}
%~ \label{sec:heuristics}
% !TEX root =  ../main.tex
%~ \subsection{Manipulability-based heuristics for contact selection}
\section{Manipulability-based heuristics for contact selection}
\label{sec:heuristics}
This Section proposes heuristics to select a contact that optimizes desired capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion,
a high velocity in a given direction, or to stay away from singular configurations.
We derive three such heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.

@@

\subsubsection{Manipulability-based heuristics}
From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.

@@

If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
\begin{equation}
h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
\end{equation}

$h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:

>>>>

%~ \section{Heuristics and criteria for contact selection}
%~ \label{sec:heuristics}
% !TEX root =  ../main.tex
%~ \subsection{Manipulability-based heuristics for contact selection}
\section{Manipulability-based heuristics for contact selection}
\label{sec:heuristics}
This Section proposes two heuristics to select a contact that optimizes desired capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.

@@

\subsubsection{Manipulability-based heuristics}
From these definitions, we can derive two useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.

@@

%~ If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
%~ \begin{equation}
%~ h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
%~ \end{equation}

%~ $h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:
